---
title: "Pre-requisites"
author: "Andrew Parnell and Doug McNeall"
output: html_document
---

In preparation for the course please install the following, preferably in the below suggested order.

Remember you will need to bring your own personal computer for the duration of the course.

### Step 1

Install the following using the corresponding links

-	R: [http://www.r-project.org](http://www.r-project.org)

-	Rstudio (optional but recommended): [https://www.rstudio.com](https://www.rstudio.com)

-	JAGS: [http://mcmc-jags.sourceforge.net](http://sourceforge.net/projects/mcmc-jags/files/)

### Step 2

Now install the following packages by going into Rstudio (or R) and typing:
```{r,eval=FALSE}
install.packages('rjags')
```

### Step 3

Check that everything works by going to a command line terminal and typing `jags`. If it opens successfully type `exit`. 

Now go into R and copy and paste the following text:
```{r,eval=FALSE}
library(rjags)
modelstring="
model {
  y ~ dnorm(mu,1)
  mu ~ dnorm(0,1)
}
"
model=jags.model(textConnection(modelstring),data=list(y=1))
output = coda.samples(model,n.iter=1000,variable.names='mu') # Burnin period
plot(output)
```
This should produce a two-panel plot.

### Troubleshooting

If you run into any problems please drop us a line at <andrew.parnell@ucd.ie> or <dougmcneall@gmail.com>.

