---
title: "Random walks and AR(p) models"
author: "Doug McNeall, Met Office"
output:
  ioslides_presentation:
    transition: slower
    widescreen: yes
  beamer_presentation: default
---

## Learning outcomes

- Understand some of the simpler time series models.
- Understand random walk processes 
- Understand AR(p) processes
- Learn how to fit some appropriate models

## Here's a time series
That's a big trend, right?

```{r random trend, echo=FALSE}
set.seed(123)
x <- cumsum(sample(c(-1, 1), size=500, replace=TRUE))
plot(x, type = 'l')
```
well, ...

## Here's how we generate the data
```{r simulate random walk, echo=TRUE}
set.seed(123)
x <- cumsum(sample(c(-1, 1), size=1000, replace=TRUE))
```
And what happens when we continue the series...
```{r, echo=FALSE}
plot(x, type = 'l')
```

## Generating a one-dimensional random walk

1. Start at zero
2. Flip a coin and move (+1) for heads (-1) for tails
3. Repeat

Where do you end up?
- Expected mean is 0 as number of steps gets large
$$ \hat{Y}_{t} = Y_{t-1} $$
- Variance gets larger with number of steps.

## Random walk applications

- Stockmarket and the efficient market hypothesis
- Levy flight (distributions of step length)
- Brownian motion and physics

## How do we know that climate change is not a random walk?

- A statistical model is just that, a convenient model of the process, not the process itself.
- The importance of process knowledge in setting up tests.
- For example, a forced temperature change can show long term persistence. Of course!

## Autoregressive models

- Output depends **linearly** on its previous values **and** a stochastic term. 

- An AR(p) process can be modelled
$$Y_{t} = \alpha + \sum_{i=1}^{p} \phi_{i} Y_{t-i} + \epsilon_{t}$$

Where $\alpha$ is a constant, $\phi_{i}$ are the model parameters and $\epsilon_{t}$ is white noise.

For example, an AR(1) process can be written
$$Y_{t} = \alpha + \phi Y_{t-1} + \epsilon_{t}$$

And an AR(2)
$$Y_{t} = \alpha + \phi_{1} Y_{t-1} + \phi_2 Y_{t-2} + \epsilon_{t}$$

## Some features of an AR(1) process

- In an AR(1) process, a shock will have an effect for an infinite amount of time!
- Although in practice, this decays if $\phi<1$.

## Simulating from an AR(1) process

```{r simulate AR1, echo=TRUE}
# Some R code to simulate data from an AR1 model
set.seed(123)
T = 100
t_seq = 1:T
sigma = 1
alpha = 1
phi = 0.6
y = rep(NA,T)
y[1] = rnorm(1,0,sigma)
for(t in 2:T) y[t] = rnorm(1,alpha + phi * y[t-1], sigma)
```

## Increasing $\phi$ parameter
```{r plot phi, echo=FALSE}
ar1sim <- function(T=200, sigma=0.5, alpha=0, phi=0.6){
  set.seed(123)
  y = rep(NA,T)
  y[1] = rnorm(1,0,sigma)
  for(t in 2:T) y[t] = rnorm(1,alpha + phi*y[t-1], sigma)
  y
}
par(mfrow = c(3,1), mar = c(2,4,0,1))
plot(ar1sim(phi=0.1),type='l', ylim = c(-3,3))
plot(ar1sim(phi=0.6),type='l', ylim = c(-3,3))
plot(ar1sim(phi=0.9),type='l', ylim = c(-3,3))

```

## Simulating from an AR(p) process

```{r simulate ARp, echo=FALSE}
# Also simulate an AR(p) process
p = 3
phi2 = c(0.5,0.1,-0.02)
y2 = rep(NA,T)
y2[1:p] = rnorm(p,0,sigma)
for(t in (p+1):T) y2[t] = rnorm(1,alpha + sum( phi2 * y2[(t-1):(t-p)] ), sigma)
plot(t_seq,y2,type='l')
```



